Global code motion of congruent instructions

### 2017-04-18 paper.tex
add comment about 15% perf uplift on motivating example

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
We present a \gcm{} (\GCM{}) compiler optimization which schedules
congruent instructions across the program.  Not only this saves code size, it
exposes redundancies in some cases, it exposes more instruction level
parallelism in the basic block to which instructions are moved, and it enables
other passes like loop invariant motion to remove more redundancies. The cost
model to drive the code motion is based on live range analysis on \SSA{}
representation such that the (virtual) register pressure does not increase, which
translates to 2\% fewer spills while compiling \SPEC{} benchmark suite.

@@

\section{Introduction}
\label{sec:intro}
Compiler techniques to remove redundant computations are composed of an analysis
phase that detects identical computations in the program and a transformation
phase that reduces the number of run-time computations.  Classical scalar
optimizations like CSE \cite{dragonbook} work very well on single basic blocks but
when it comes to detect redundancies across basic blocks these techniques fall
short: more complex passes like GCSE and \PRE{} have been designed to handle these
cases based on dataflow analysis \cite{morel1979global}.  At first these
techniques were described in the classical data-flow analysis framework, and
later the use of the \SSA{} representation lowered the cost in terms of compilation
time \cite{briggs1994effective,chow1997new,kennedy1999partial} and brought these
techniques in the main stream: nowadays \SSA{} based \PRE{} is available in every
industrial compiler.

This paper describes \gcm{} of congruent instructions \cite{briggs1997}, a
technique that uses the information computed for \PRE{} to detect identical
computations but has a transformation phase whose goal differs from \PRE{}: it
removes identical computations from different branches of execution.  These
identical computations in different branches of execution are not redundant
computations at run-time and the number of run-time computations is not
reduced. It is not a redundancy elimination pass, and thus it has different cost
function and heuristics than \PRE{} or CSE. It is more similar to global code
scheduling \cite{dragonbook,click1995global} in the sense that it will only move
computations. Code hoisting, for example, can reduce the critical path length of
execution in \ooo{} machines. As more instructions are available at the
hoisting point, the hardware has more independent instructions to reorder. The
following example illustrates how hoisting can improve performance by exposing
more ILP.

\begin{verbatim}
float fun(float d, float min, float max, float a) {
  float tmin, tmax, inv;

  inv = 1.0f / d;
  if (inv >= 0) { tmin = (min - a) * inv; tmax = (max - a) * inv; }
  else          { tmin = (max - a) * inv; tmax = (min - a) * inv; }
  return tmax + tmin;
}
\end{verbatim}

@@

Now if the computation of tmax and tmin are hoisted outside the
conditionals, the C code version would look like this:
\begin{verbatim}
float fun(float d, float min, float max, float a) {
  float inv = 1.0f / d;
  float x = (min - a) * inv;
  float y = (max - a) * inv;
  float tmin = x, tmax = y;
  if (inv < 0) { tmin = y; tmax = x; }
  return tmax + tmin;
}

\end{verbatim}

In this code the division operation can be executed in parallel with the two
subtractions because there are no dependencies among them. So the total number
of cycles will be $max(C_{div}, C_{sub}) + C_{mul} = C_{div} + C_{mul}$; since
$C_{div}$ is usually much greater than $C_{sub}$ \cite{x86,aarch64}.

%GCC: 23286 has interesting test cases.

In fact, there are several advantages of \GCM{}.
\begin{itemize}
\item it helps reduce the code size of the program by replacing multiple
  instructions with one although the final code size may be higher because of
  improved inlining heuristics; functions become cheaper to inline when their
  code size decreases because the inliner heuristics depend on instruction
  count;
\item it exposes more instruction level parallelism to the later compiler
  passes. By hoisting identical computations to be executed earlier, instruction
  schedulers can move heavy computations earlier in order to avoid pipeline
  bubbles;
\item it reduces branch misprediction penalty on \ooo{} processors with
  speculative execution of branches: by hoisting or sinking expressions out of
  branches, it can effectively reduce the amount of code to be speculatively
  executed and hence reduce the critical path;
\item it reduces interference or register pressure with an appropriate
  cost-model: we have used a cost model in our implementation which only hoists
  or sinks when the register pressure is reduced;
\item it reduces the total number of instructions to be executed for SIMD
  architectures which execute all code in branches based on masking or
  predication;
\item it may improve loop vectorization by reducing a loop with control flow to
  a loop with a single BB, should all the instructions in a conditional get
  hoisted or sinked;
\item it enables more loop invariant code motion (LICM): as LICM passes, in
  general, cannot effectively reason about instructions within conditional
  branches in the context of loops, code-hoisting is needed to move instructions
  out of conditional expressions and expose them to LICM.
\end{itemize}

@@

There are a lot of bug reports in \GCC{} and \LLVM{} bugzillas
\cite{GCCCodeHoistingBugs, LLVMCodeHoistingBugs}, showing the interest in having
a more powerful code hoist transform.  The current \LLVM{} implementation of
code hoisting in SimplifyCFG.cpp is very limited to hoisting from identical
basic blocks: the instructions of two sibling basic blocks are read at the same
time, and all the instructions of the blocks are hoisted to the common parent
block as long as the compiler is able to prove that the instructions are
equivalent.  This implementation does not allow for an easy extension: first in
terms of compilation time overhead the implementation is quadratic in number of
instructions to bisimulate and second, the equivalence of instructions is
computed by comparing the operands which is neither general nor scalabale.

\citet{dhamdhere1988fast}, \citet{steven1997advanced} mention
code hoisting in a data flow framework. A list of Very Busy Expressions (VBE)
computed which are hoisted in a basic block where the expression is
anticipable (all the operands are available). This algorithm would hoist as far
as possible without regarding the impact on register pressure and as such a cost
model will be required.  Also the description of VBE is based on the classic
dataflow model and an adaptation to a sparse \SSA{} representation is required.

@@

\GCC{} recently got code-hoisting \cite{GCCCodeHoisting} which is implemented as
part of \GVN{}-\PRE{}: it uses the set of ANTIC\_IN and AVAIL\_OUT value
expressions computed for \PRE{}. ANTIC\_IN[B] contains very busy expressions at
basic block B, i.e., values computed on all paths from B to exit and
AVAIL\_OUT[B] contains values which are already available. The algorithm hoists
top down to a predecessor.  It uses ANTIC\_IN[B] to know what expressions will
be computed on every path from B to exit, and can be computed in B.  It uses
AVAIL\_OUT[B] to subtract out those values already being computed.  The cost
function is: for each hoist candidate, if all successors of B are dominated by
B, then we know insertion into B will eliminate all the remaining computations.
It then checks to see if at least one successor of B has the value available.
This avoids hoisting it way up the chain to ANTIC\_IN[B].  It also checks to
ensure that B has multiple successors, since hoisting in a straight line is
pointless.  The algorithm continues top down the dominator tree, working in
tandem with \PRE{} until no more hoisting is possible.  One advantage of \GCC{}
implementation is that it works in sync with the \GVN{}-\PRE{} such that when
new hoisting opportunities are created by \GVN{}-\PRE{}, code-hoisting will
hoist them.

\citet{click1995global} describes aggressive \gcm{} to first
schedule all the instructions as early as possible. This results in very long
live ranges which is mitigated by again scheduling all the instructions as late
as possible. They report sppeedup of as high as $23\%$.

@@

\subsection{Liveness using DJ-graph}
DJ-graph \cite{Sreedhar1996} is a data structure that augments the dominator
tree with join-edges to keep track of data-flow in a program. We use DJ-graph
and merge-sets to compute liveness of variables as illustrated in
\cite{das2012}.  It is very efficient for computing liveness and does not
require any bitvectors to be maintained for each basic block. The underlying
simplicity of liveness computation is due to \SSA{} form, where the the values
only flow (from def to use) either through dominator edges or the join edges
(where we insert a PHI). The DJ-graph contains both these edges which allows for
computation of merge-sets for each basic block, i.e., a set of all basic blocks
where the values can flow from a particular basic block. We have implemented the
merge-set computation based on a DJ-graph as illustrated in \cite{das2005}. A
simplified version of the merge-set that we implemented is presented here:

@@

We compute merge sets, of the control flow graph, which remain same throughout
the \gcm{} transformation. For \gcm{} we only want to know if a use
operand is a kill (to compute changes in register pressure.) For that we only
need to know whether the use is also required later in the execution path. A
variable is live out of a basic block B if it is used in the merge set of B. We
compute the live-out relation on-demand when profitability of hoistable/sinkable
candidates is to be evaluated. A simplified version of
\emph{isLiveOutUsingMergeSet} is presented here:

\begin{verbatim}
// Return true if variable A is liveout from basic block N
bool isLiveOutUsingMergeSet(Node N, Variable A)
  if (A is defined in N)
    if (A is used outside any basic block other than N)
      return true
    return false

@@

The original algorithm presented in \cite{das2012} has mismatched types in terms
of uses and nodes because each node can have many instructions and hence many
uses. Also, while iterating on the dominator of each user of a variable we may
reach to the beginning of a function, in that case the inner while loop needs to
terminate. These two cases were missing from the algorithm and we came across
them during implementation. If a variable is not live-out of a basic block, it
still may be used later in the basic block: to establish whether a use of a
variable is a kill we iterate on all the subsequent instructions in a basic
block checking for uses, the \emph{isKill} function uses
\emph{isLiveOutUsingMergeSet} to find out if an operand is killed in an
instruction.

@@

\subsection{Finding candidates to move}
\label{subsec:finding-candidates}
The first step is to find a set of congruent instructions
\cite{briggs1997}. This is performed by a linear scan of all instructions of the
program and classifying them by their value numbers. We could compute available
and anticipable sets as computed by \GCC{'}s code-hoisting but that would be a lot
of data structures to maintain at each basic block level. For \GCC{} it makes sense
because their code-hoisting is integrated with \GVN{}-\PRE{} which already has those
data structures available.

@@

\subsubsection{Barriers}
Barriers are based on the concept of pinned instructions \cite{click1995global}
but extended to adapt to \LLVM{} IR. Since a basic block in \LLVM{} IR is
actually an extended basic block because there might be calls in the middle of a
basic block which might not return. Essentially, any instruction that cannot
guarantee progress is marked as a barrier. In the absence of context, as in our
current implementation, some instructions which might be safe are still
classified as barriers, e.g., volatile loads/stores, calls with missing
attributes.

\begin{verbatim}
computing barriers:
for each basic block B in a function:
  barrier_found = false
  for each instruction I in B:
    if I does not guarantee progress:
      mark I as a barrier instruction
      barrier_found = true
      break;

@@

\begin{verbatim}
computing downward-safety of hoistable instructions at hoist-point
  Worklist = list of all basic blocks of hoistable instructions
  for each basic block B in the dominator tree starting at hoist-point:
    if Worklist is empty, return false // Path exists! not downward safe
    if B is in Worklist:
      remove B from worklist
      remove subtree with root at B // Available at B => downward safe from B
    if B is a leaf basic block, return false  // Path exists! not downward safe
    if B dominates hoist-point, return false  // Back edge
  return true
\end{verbatim}

@@

There is a special case for hoisting load instructions when the hoist-point is
the predecessor basic block for all the loads. Even if the register pressure
would increase, we prefer to hoist loads in this case. The reason being, that
would make the loaded value available by the time it is used. Also, because
stores and calls are hoisted the least, see Table~\ref{tab:code-motion-metric},
the performance does not change much whether they are hoisted or not.

@@

\GCM{} basically consists of two parts, i.e., hoisting and sinking. This
implementation only moves congruent instructions. An immediate guarantee of this
approach is gain in code size (the final executable may be of larger size
because of more inlining). The algorithm prefers hoisting to sinking. If the
dependency of a hoistable candidate is in the same basic block as the candidate,
then the dependency must also be hoistable, otherwise hoisting will be illegal or
would require a complicated code generation to make it legal. The current
algorithm discards cases if the dependency is neither hoistable nor
rematerializable.

>>>>

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
We present a \gcm{} (\GCM{}) compiler optimization which schedules
congruent instructions across the program.  Not only this saves code size, it
exposes redundancies in some cases, it exposes more instruction level
parallelism in the basic-block to which instructions are moved, and it enables
other passes like loop invariant motion to remove more redundancies. The cost
model to drive the code motion is based on live range analysis on \SSA{}
representation such that the (virtual) register pressure does not increase, which
translates to 2\% fewer spills while compiling \SPEC{} benchmark suite.

@@

\section{Introduction}
\label{sec:intro}
Compiler techniques to remove redundant computations are composed of an analysis
phase that detects identical computations in the program and a transformation
phase that reduces the number of run-time computations.  Classical scalar
optimizations like CSE \cite{dragonbook} work very well on single basic-blocks but
when it comes to detect redundancies across basic-blocks these techniques fall
short: more complex passes like GCSE and \PRE{} have been designed to handle these
cases based on dataflow analysis \cite{morel1979global}.  At first these
techniques were described in the classical data-flow analysis framework, and
later the use of the \SSA{} representation lowered the cost in terms of compilation
time \cite{briggs1994effective,chow1997new,kennedy1999partial} and brought these
techniques in the main stream: nowadays \SSA{} based \PRE{} is available in every
industrial compiler.

This paper describes \gcm{} (\GCM{}) of congruent instructions \cite{briggs1997}, a
technique that uses the information computed for \PRE{} to detect identical
computations but has a transformation phase whose goal differs from \PRE{}: it
removes identical computations from different branches of execution.  These
identical computations in different branches of execution are not redundant
computations at run-time and the number of run-time computations is not
reduced. It is not a redundancy elimination pass, and thus it has different cost
function and heuristics than \PRE{} or CSE. It is more similar to global code
scheduling \cite{dragonbook,click1995global} in the sense that it will only move
computations. Code hoisting, for example, can reduce the critical path length of
execution in \ooo{} machines. As more instructions are available at the
hoisting point, the hardware has more independent instructions to reorder. The
following example illustrates how hoisting can improve performance by exposing
more ILP.

\begin{verbatim}
float fun(float d, float min, float max, float a) {
  float tmin, tmax;
  float inv = 1.0f / d;
  if (inv >= 0) { tmin = (min - a) * inv; tmax = (max - a) * inv; }
  else          { tmin = (max - a) * inv; tmax = (min - a) * inv; }
  return tmax + tmin;
}
\end{verbatim}

@@

Now if the computation of tmax and tmin are hoisted outside the
conditionals, the C code version would look like this:
\begin{verbatim}
float fun(float d, float min, float max, float a) {
  float inv = 1.0f / d;
  float x = (min - a) * inv;
  float y = (max - a) * inv;
  float tmin = x, tmax = y;
  if (inv < 0) { tmin = y; tmax = x; }
  return tmax + tmin;
}
\end{verbatim}

In this code the division operation can be executed in parallel with the two
subtractions because there are no dependencies among them. So the total number
of cycles will be $max(C_{div}, C_{sub}) + C_{mul} = C_{div} + C_{mul}$; since
$C_{div}$ is usually much greater than $C_{sub}$ \cite{x86,aarch64}. We have
seen the performance of an \ooo{} processor going up by $15\%$ on a benchmark
containing the above motivating example running in a tight loop.  In fact, there
are several advantages of \GCM{}:
\begin{itemize}
\item it helps reduce the code size of the program by replacing multiple
  instructions with one although the final code size may be higher because of
  improved inlining heuristics; functions become cheaper to inline when their
  code size decreases because the inliner heuristics depend on instruction
  count;
\item it exposes more instruction level parallelism to the later compiler
  passes. By hoisting identical computations to be executed earlier, instruction
  schedulers can move heavy computations earlier in order to avoid pipeline
  bubbles;
\item it reduces branch misprediction penalty on \ooo{} processors with
  speculative execution of branches: by hoisting or sinking expressions out of
  branches, it can effectively reduce the amount of code to be speculatively
  executed and hence reduce the critical path;
\item it reduces interference or register pressure with an appropriate
  cost-model: we have used a cost model in our implementation which only hoists
  or sinks when the register pressure is reduced;
\item it reduces the total number of instructions to be executed for SIMD
  architectures which execute all code in branches based on masking or
  predication;
\item it may improve loop vectorization by reducing a loop with control flow to
  a loop with a single basic-block, should all the instructions in a conditional get
  hoisted or sinked;
\item it enables more loop invariant code motion (LICM): as LICM passes, in
  general, cannot effectively reason about instructions within conditional
  branches in the context of loops, code-hoisting is needed to move instructions
  out of conditional expressions and expose them to LICM.
\end{itemize}

@@

There are a lot of bug reports in \GCC{} and \LLVM{} bugzillas
\cite{GCCCodeHoistingBugs, LLVMCodeHoistingBugs}, showing the interest in having
a more powerful code hoist transform.  The current \LLVM{} implementation of
code hoisting in SimplifyCFG.cpp is very limited to hoisting from identical
basic-blocks: the instructions of two sibling basic-blocks are read at the same
time, and all the instructions of the blocks are hoisted to the common parent
block as long as the compiler is able to prove that the instructions are
equivalent.  This implementation does not allow for an easy extension: first in
terms of compilation time overhead the implementation is quadratic in number of
instructions to bisimulate and second, the equivalence of instructions is
computed by comparing the operands which is neither general nor scalabale.

\citet{dhamdhere1988fast}, \citet{steven1997advanced} mention
code hoisting in a data flow framework. A list of Very Busy Expressions (VBE)
computed which are hoisted in a basic-block where the expression is
anticipable (all the operands are available). This algorithm would hoist as far
as possible without regarding the impact on register pressure and as such a cost
model will be required.  Also the description of VBE is based on the classic
dataflow model and an adaptation to a sparse \SSA{} representation is required.

@@

\GCC{} recently got code-hoisting \cite{GCCCodeHoisting} which is implemented as
part of \GVN{}-\PRE{}: it uses the set of ANTIC\_IN and AVAIL\_OUT value
expressions computed for \PRE{}. ANTIC\_IN[B] contains very busy expressions at
basic-block B, i.e., values computed on all paths from B to exit and
AVAIL\_OUT[B] contains values which are already available. The algorithm hoists
top down to a predecessor.  It uses ANTIC\_IN[B] to know what expressions will
be computed on every path from B to exit, and can be computed in B.  It uses
AVAIL\_OUT[B] to subtract out those values already being computed.  The cost
function is: for each hoist candidate, if all successors of B are dominated by
B, then we know insertion into B will eliminate all the remaining computations.
It then checks to see if at least one successor of B has the value available.
This avoids hoisting it way up the chain to ANTIC\_IN[B].  It also checks to
ensure that B has multiple successors, since hoisting in a straight line is
pointless.  The algorithm continues top down the dominator tree, working in
tandem with \PRE{} until no more hoisting is possible.  One advantage of \GCC{}
implementation is that it works in sync with the \GVN{}-\PRE{} such that when
new hoisting opportunities are created by \GVN{}-\PRE{}, code-hoisting will
hoist them.

\citet{click1995global} describes aggressive \gcm{} to first
schedule all the instructions as early as possible. This results in very long
live ranges which is mitigated by again scheduling all the instructions as late
as possible. They report a speedup of as high as $23\%$.

@@

\subsection{Liveness using DJ-graph}
DJ-graph \cite{Sreedhar1996} is a data structure that augments the dominator
tree with join-edges to keep track of data-flow in a program. We use DJ-graph
and merge-sets to compute liveness of variables as illustrated in
\cite{das2012}.  It is very efficient for computing liveness and does not
require any bitvectors to be maintained for each basic-block. The underlying
simplicity of liveness computation is due to \SSA{} form, where the the values
only flow (from def to use) either through dominator edges or the join edges
(where we insert a PHI). The DJ-graph contains both these edges which allows for
computation of merge-sets for each basic-block, i.e., a set of all basic-blocks
where the values can flow from a particular basic-block. We have implemented the
merge-set computation based on a DJ-graph as illustrated in \cite{das2005}. A
simplified version of the merge-set that we implemented is presented here:

@@

We compute merge sets, of the control flow graph, which remain same throughout
the \gcm{} transformation. For \gcm{} we only want to know if a use
operand is a kill (to compute changes in register pressure.) For that we only
need to know whether the use is also required later in the execution path. A
variable is live out of a basic-block B if it is used in the merge set of B. We
compute the live-out relation on-demand when profitability of hoistable/sinkable
candidates is to be evaluated. A simplified version of
\emph{isLiveOutUsingMergeSet} is presented here:

\begin{verbatim}
// Return true if variable A is liveout from basic-block N
bool isLiveOutUsingMergeSet(Node N, Variable A)
  if (A is defined in N)
    if (A is used outside any basic-block other than N)
      return true
    return false

@@

The original algorithm presented in \cite{das2012} has mismatched types in terms
of uses and nodes because each node can have many instructions and hence many
uses. Also, while iterating on the dominator of each user of a variable we may
reach to the beginning of a function, in that case the inner while loop needs to
terminate. These two cases were missing from the algorithm and we came across
them during implementation. If a variable is not live-out of a basic-block, it
still may be used later in the basic-block: to establish whether a use of a
variable is a kill we iterate on all the subsequent instructions in a
basic-block checking for uses, the \emph{isKill} function uses
\emph{isLiveOutUsingMergeSet} to find out if an operand is killed in an
instruction.

@@

\subsection{Finding candidates to move}
\label{subsec:finding-candidates}
The first step is to find a set of congruent instructions
\cite{briggs1997}. This is performed by a linear scan of all instructions of the
program and classifying them by their value numbers. We could compute available
and anticipable sets as computed by \GCC{'}s code-hoisting but that would be a lot
of data structures to maintain at each basic-block level. For \GCC{} it makes sense
because their code-hoisting is integrated with \GVN{}-\PRE{} which already has those
data structures available.

@@

\subsubsection{Barriers}
Barriers are based on the concept of pinned instructions \cite{click1995global}
but extended to adapt to \LLVM{} IR. Since a basic-block in \LLVM{} IR is
actually an extended basic-block because there might be calls in the middle of a
basic-block which might not return. Essentially, any instruction that cannot
guarantee progress is marked as a barrier. In the absence of context, as in our
current implementation, some instructions which might be safe are still
classified as barriers, e.g., volatile loads/stores, calls with missing
attributes.

\begin{verbatim}
computing barriers:
for each basic-block B in a function:
  barrier_found = false
  for each instruction I in B:
    if I does not guarantee progress:
      mark I as a barrier instruction
      barrier_found = true
      break;

@@

\begin{verbatim}
computing downward-safety of hoistable instructions at hoist-point
  Worklist = list of all basic-blocks of hoistable instructions
  for each basic-block B in the dominator tree starting at hoist-point:
    if Worklist is empty, return false // Path exists! not downward safe
    if B is in Worklist:
      remove B from worklist
      remove subtree with root at B // Available at B => downward safe from B
    if B is a leaf basic-block, return false  // Path exists! not downward safe
    if B dominates hoist-point, return false  // Back edge
  return true
\end{verbatim}

@@

There is a special case for hoisting load instructions when the hoist-point is
the predecessor basic-block for all the loads. Even if the register pressure
would increase, we prefer to hoist loads in this case. The reason being, that
would make the loaded value available by the time it is used. Also, because
stores and calls are hoisted the least, see Table~\ref{tab:code-motion-metric},
the performance does not change much whether they are hoisted or not.

@@

\GCM{} basically consists of two parts, i.e., hoisting and sinking. This
implementation only moves congruent instructions. An immediate guarantee of this
approach is gain in code size (the final executable may be of larger size
because of more inlining). The algorithm prefers hoisting to sinking. If the
dependency of a hoistable candidate is in the same basic-block as the candidate,
then the dependency must also be hoistable, otherwise hoisting will be illegal or
would require a complicated code generation to make it legal. The current
algorithm discards cases if the dependency is neither hoistable nor
rematerializable.
